taxize - taxonomic search and retrieval in R

### 2013-06-21 inst/doc/taxize_withcode_plos.Rnw
minor changes
removed comments from code blocks, carls changes incorporated, some of owens

\section*{Introduction}
Evolution by natural selection has led to a hierarchical relationship among all living organisms.  Thus, species are categorized using a taxonomic hierarchy, starting with the binomial species name (e.g, \emph{Homo sapiens}), moving up to genus (\emph{Homo}), then family (\emph{Hominidae}), and on up to Domain (\emph{Eukarya}). Biologists, whether studying organisms at the cell, organismal, or community level, can put their study taxa into taxonomic context, allowing them to know close and distant relatives, find relevant literature, and more. Discovering the correct taxonomic names is, unfortunately, not straightforward. Taxonomic names often change due to name changes at the generic or specific levels, lumping or splitting lower taxa (genera, species) among higher taxa (families), and name spelling changes. In addition, there is no one authoritative taxonomic names source. Instead, there are essentially competing sources (e.g., uBio, Tropicos, ITIS) that may have different accepted names for the same taxon. The goal of taxize, an R package in development, is to make all use cases having to do with retrieving and resolving taxonomic names easy and replicable.

@@

In taxize, we have written a suite of R functions that interact with many taxonomic data sources via their web APIs (Table~\ref{tab:a}). The interface to each function is usually a simple list of species names, just as a user would do with a web API. Therefore, we hope moving from a web to R interface for taxonomic names will be relatively seamless (if one is already nominally familiar with R).

@@

All of these tools provide great ways to search for taxonomic names and resolve them in some cases. However, scientists ideally need a tool that can be used programmatically, and thereby facilitate reproducible research. The goal of taxize is to make it easy to create reproducible and easy to use workflows for searching for taxonomic names, resolving them, getting higher taxonomic names, and other tasks related to research dealing with species.

@@

\section*{Data sources}
taxize uses many data sources (Table~\ref{tab:a}), and more can easily be added. There are two common tasks provided by the data sources: name search and name resolution. Other functionality in taxize includes retrieving a classification tree for a species, or retrieving child taxa of a focal taxon. One of the data sources (Phylomat) returns phylogenies, while another (NCBI) returns genetic sequence data. However, there are other R packages that are focused solely on sequence data, such as rsnps \cite{chamberlain2013}, rentrez \cite{winter2013}, BoSSA \cite{lefeuvre2010}, and ape \cite{paradis2004}, so taxize will not venture deeply into these other domains.

Some of the data sources taxize interacts with require authentication. That is, in addition to the search terms the user provides (e.g., \emph{Homo sapiens}), the data provider requires an alphanumeric identification key so that they can better manage their servers, collect analytics, and shut down users that abuse the API. The services that do require an API key in taxize are: Encyclopedia of Life (EOL), the Universal Biological Indexer and Organizer (uBio), Tropicos, and Plantminer. You can easily obtain an API key by visiting the website of each service (see Table~\ref{tab:a} for links to each site). There are two ways of using your API keys. First, you can pass in your API key in a function call (e.g., \emph{ubio\_namebank(srchName='Ursus americanus', key='your\_alphanumeric\_key')}). Second, you can store your API keys in your .Rprofile file. On a Mac this file is at \emph{/yourhomefolder/.Rprofile}; on a Windows machine at \emph{/yourhomefolder/.Rprofile}; and on Linux at \emph{/yourhomefolder/.Rprofile}. This is an hidden file, so open up this file in your terminal (e.g., \emph{open .Rprofile}), and add the API key as an entry like \emph{options(myapikey = 'your\_alphanumeric\_key')}. We recommend the second option as it simplifies function calls.

@@

@@

<<installtaxize, eval=TRUE>>=
## From CRAN
install.packages("taxize")

## From GitHub
install.packages("devtools")
require(devtools)
install_github("taxize_", "ropensci")
@

Loading taxize into your R session

<<loadtaxize, eval=TRUE>>=
library(taxize)
@

@@

This is a common task in biology. We often have a list of species names and we want to know if a) we have the most up to date names, b) our names are spelled correctly, and c) if we have common names, we likely need the scientific names. One way to resolve names is via the Global Names Resolver (GNR) service provided by the Encyclopedia of Life \cite{eolgnr}.

<<resolvenames, eval=TRUE, cache=TRUE>>=
# Here, we are searching for two misspelled names
temp <- gnr_resolve(names = c("Helianthos annus", "Homo saapiens"), returndf = TRUE)

# Let's take a peek at the data, excluding the data source ID and score columns
temp[,-c(1,4)]
@

It looks like the correct spellings are \emph{Helianthus annuus} and \emph{Homo sapiens}. Another approach uses the Taxonomic Name Resolution Service via the Taxosaurus API \cite{taxosaurus} developed by iPLant and the Phylotastic organization.

<<tnrs, eval=TRUE,cache=TRUE>>=
# A list of species names, some of which are misspelled
mynames <- c("Helianthus annuus", "Pinus contort", "Poa anua", "Abis magnifica",
  	"Rosa california", "Festuca arundinace", "Sorbus occidentalos","Madia sateva")

# And we'll call the API with the tnrs function, and remove a few columns
tnrs(query = mynames)[,-c(5:7)]
@

It looks like there are a few corrections: e.g., \emph{Madia sateva} should be \emph{Madia sativa}, and \emph{Rosa california} should be \emph{Rosa californica}. Note that this search worked because fuzzy matching was employed to retrieve names that were close, but not exact matches. Fuzzy matching is only available for plants in the TNRS service, so we advise using EOL's Global Names Resolver if you need to resolve animal names.

@@

<<itisclass, eval=TRUE,cache=TRUE>>=
specieslist <- c("Abies procera","Pinus contorta")
classification(get_tsn(specieslist, "sciname"))
@

@@

Instead of a full classification, you may only want a single name, say a family name for your species of interest. The function \emph{tax\_name} is built just for this purpose. And you can specify the data source you retrieve the taxonomic name from with the \emph{db} parameter.

<<taxname, eval=TRUE,cache=TRUE>>=
# Using ITIS 
tax_name(query="Helianthus annuus", get="family", db="itis")

# Using NCBI
tax_name(query="Helianthus annuus", get="family", db="ncbi")
@

\subsection*{Interactive name selection}
In the previous section we used the function \emph{get\_tsn} to get a classification for two plant species. Below are a few examples. When you run these examples in R, you are presented with a command prompt asking for the row that contains the name you would like back; that output is not printed below fro brevity.

<<interactive, eval=TRUE, echo=TRUE, cache=TRUE>>=
# In this example, this search term has many matches. The function returns a data.frame of the matches, and asks for user input. Enter the row number of the name you want. 
get_tsn(searchterm="Elisabethae", searchtype="sciname")

# You can pass in a long list of taxonomic names
splist <- c("annona cherimola", 'annona muricata', "quercus robur",
  	"shorea robusta", "pandanus patina", "oryza sativa", "durio zibethinus")
get_tsn(searchterm=splist, searchtype="sciname")

# In this example, note that no match at all returns an NA
get_uid(sciname=c("Chironomus riparius", "aaa vva"))
@

@@

Behind the scenes the function \emph{phylomatic\_tree} retrieves a Taxonomic Serial Number (TSN) from ITIS for each species name, then a string is created for each species like this \emph{poaceae/oryza/oryza\_sativa} (with format 'family/genus/genus\_epithet'). These strings are submitted to the Phylomatic API, and if no errors occur, a phylogeny in newick format is returned. The \emph{phylomatic\_tree()} function also cleans up the newick string and converts it to an \textbf{ape} \emph{phylo} object. The output from \emph{phylomatic\_tree()} is a \emph{phylo} object, which can be used for plotting and phylogenetic analyses.

@@

<<downstream, eval=TRUE, cache=TRUE>>=
# Search for all species downstream from Apis
col_downstream(name="Apis", downto="Species")[[1]]
@

@@

<<iucn, eval=TRUE, cache=TRUE>>=
ia <- iucn_summary(c("Panthera uncia", "Lynx lynx"))
iucn_status(ia)
@

@@

<<genbank, eval=TRUE, cache=TRUE>>=
# Serch for available genes in GenBank.
out <- get_genes_avail(taxon_name="Umbra limi", seqrange = "1:2000", getrelated=FALSE)

# Get a list of available genes, removing non-unique records.
# unique(out$genesavail)

# Does the string 'RAG1' exist in any of the gene names?
out[grep("RAG1", out$genesavail, ignore.case=T),]
@

@@

To analyse trait-composition abundance data must be matched with trait databases like \cite{usseglio-polatera_biological_2000}. However these two data tables may contain species information on different taxonomic levels and perhabs data must be aggregated to a joint taxomic level.

@@

\begin{table}[!ht]
\caption{
\bf{Data sources used in taxize, tasks available, and links to them}}
\begin{tabular}{|l|c|c|c|c|}
\hline
Source name & Name search & Name resolution & Phylogeny & Sequences  \\
\hline
Encyclopedia of Life \cite{eol} & Yes & See GNR below & No & No \\
Integrated Taxonomic Information System \cite{itis} & Yes & Synonyms & No & No \\
iPlant Taxonomic Name Resolution Service \cite{tnrs} & Yes & Yes & No & No \\
Phylomatic \cite{webb2005} & No & No & Yes & No \\
uBio \cite{ubio} & Yes & Yes & No & No \\
Global Names Resolver \cite{eolgnr} & Yes & Yes & No & No \\
Global Names Index \cite{eolgnr} & Yes & No & No & No \\
IUCN Red List \cite{iucn} & Yes & No & No & No \\
Tropicos \cite{tropicos} & Yes & Yes & No & No \\
Plantminer \cite{carvalho2010plantminer} & Yes & No & No & No \\
The Plant List & Yes & Yes & No & No \\
Catalogue of Life \cite{col} & Yes & Yes & No & No \\
National Center for Biotechnology Information \cite{ncbi} & Yes & X & Yes \textsuperscript{a} & Yes \\
\hline
\end{tabular}
\begin{flushleft}\textsuperscript{a} Web only, no API, see \url{http://1.usa.gov/11R446a}
\end{flushleft}
\label{tab:a}
\end{table}

>>>>

\section*{Introduction}
Evolution by natural selection has led to a hierarchical relationship among all living organisms.  Thus, species are categorized using a taxonomic hierarchy, starting with the binomial species name (e.g, \emph{Homo sapiens}), moving up to genus (\emph{Homo}), then family (\emph{Hominidae}), and on up to Domain (\emph{Eukarya}). Biologists, whether studying organisms at the cell, organismal, or community level, can put their study taxa into taxonomic context, allowing them to know close and distant relatives, find relevant literature, and more.

The use of taxonomic names is, unfortunately, not straightforward. Taxonomic names often change due to name changes at the generic or specific levels, lumping or splitting lower taxa (genera, species) among higher taxa (families), and name spelling changes. For example, a study found that a compilation of 308,000 plant observations from 51 digitized herbarium records had 22,100 unique taxon names, of which only ~13,000 were accepted names \cite{weiser2007,boyle2013}. The scale of this study may be larger than most, but the problem persists for any size study. In addition, there is no one authoritative taxonomic names source for all taxa - although, there are taxon specific sources that many scientists use that study that taxon. Different sources (e.g., uBio, Tropicos, ITIS) may have different accepted names for the same taxon. For example, while the Integrated Taxonomic Information Service (ITIS) has \emph{Helianthus x glaucus} as an accepted name, The Plant List \url{http://www.theplantlist.org} has that name as unresolved, but has \emph{Helianthus glaucus} as an accepted name, while ITIS doesn't have the name.

One attempt to help inconsistencies in taxonomy is the use of numeric codes. For example, ITIS assigns a Taxonomic Serial Number (TSN) to each taxon, while the Universal Biological Indexer and Organizer (uBio) assigns each taxon a NameBank identifier (namebankID), and Tropicos assigns their own identifier to each taxon. Codes are helpful within a database as they can easily refer to, for example, \emph{Helianthus annuus} with a code like 123456 instead of its whole name. However, each database uses their own code; in this case for \emph{Helianthus annuus}, ITIS uses 36616, uBio uses 2658020, and Tropicos uses 40022652. Yet, there are no universal codes for taxa across databases, leading to additional confusion. Last, name comparisons across databases have to be done with the actual names, not the codes.

@@

The goal of taxize, an R package in development, is to make all use cases having to do with retrieving and resolving taxonomic names easy and replicable. In taxize, we have written a suite of R functions that interact with many taxonomic data sources via their web APIs (Table~\ref{tab:a}). The interface to each function is usually a simple list of species names, just as a user would do with a web API. Therefore, we hope moving from a web to R interface for taxonomic names will be relatively seamless (if one is already nominally familiar with R).

@@

All of these tools provide ways to search for taxonomic names and resolve them in some cases. However, scientists ideally need a tool that can be used programmatically, and thereby facilitate reproducible research. The goal of taxize is to make it easy to create reproducible and easy to use workflows for searching for taxonomic names, resolving them, getting higher taxonomic names, and other tasks related to research dealing with species.

@@

\section*{Data sources and package details}
taxize uses many data sources (Table~\ref{tab:a}), and more can easily be added. There are two common tasks provided by the data sources: name search and name resolution. Other functionality in taxize includes retrieving a classification tree for a species, or retrieving child taxa of a focal taxon. One of the data sources (Phylomatic) returns phylogenies, while another (NCBI) returns genetic sequence data. However, there are other R packages that are focused solely on sequence data, such as rsnps \cite{chamberlain2013}, rentrez \cite{winter2013}, BoSSA \cite{lefeuvre2010}, and ape \cite{paradis2004}, so taxize will not venture deeply into these other domains.

Some of the data sources taxize interacts with require authentication. That is, in addition to the search terms the user provides (e.g., \emph{Homo sapiens}), the data provider requires an alphanumeric identification key so that they can better manage their servers, collect analytics, and shut down users that abuse the API. The services that do require an API key in taxize are: Encyclopedia of Life (EOL) \cite{eol}, the Universal Biological Indexer and Organizer (uBio) \cite{ubio}, Tropicos \cite{tropicos}, and Plantminer \cite{carvalho2010plantminer}. You can easily obtain an API key by visiting the website of each service (see Table~\ref{tab:a} for links to each site). There are two ways of using your API keys. First, you can pass in your API key in a function call (e.g., \emph{ubio\_namebank(srchName='Ursus americanus', key='your\_alphanumeric\_key')}). Second, you can store your API keys in your .Rprofile file. On a Mac this file is at \emph{/yourhomefolder/.Rprofile}; on a Windows machine at \emph{/yourhomefolder/.Rprofile}; and on Linux at \emph{/yourhomefolder/.Rprofile}. This is an hidden file, so open up this file in your terminal (e.g., \emph{open .Rprofile}), and add the API key as an entry like \emph{options(myapikey = 'your\_alphanumeric\_key')}. We recommend the second option as it simplifies function calls.

@@

taxize would not have been possible without the work of others. taxize uses httr \cite{httr} and RCurl \cite{rcurl} for doing calls to web APIs, XML \cite{xml} for parsing XML, RJSONIO \cite{rjsonio} for parsing JSON, and stringr \cite{stringr} and plyr \cite{plyr} for manipulating data.

@@

<<installtaxizecran, eval=FALSE>>=
install.packages("taxize")
@

<<installtaxizegithub, eval=FALSE>>=
install.packages("devtools")
require(devtools)
install_github("taxize_", "ropensci")
@

Load taxize into your R session.

<<loadtaxize, eval=FALSE>>=
library(taxize)
@

@@

This is a common task in biology. We often have a list of species names and we want to know if a) we have the most up to date names, b) our names are spelled correctly, and c) if we have common names, we likely need the scientific names. One way to resolve names is via the Global Names Resolver (GNR) service provided by the Encyclopedia of Life \cite{eolgnr}. Here, we are searching for two misspelled names:

<<resolvenames, eval=FALSE, cache=TRUE>>=
temp <- gnr_resolve(names = c("Helianthos annus", "Homo saapiens"), returndf = TRUE)
temp[,-c(1,4)]
@

The correct spellings are \emph{Helianthus annuus} and \emph{Homo sapiens}. Another approach uses the Taxonomic Name Resolution Service via the Taxosaurus API \cite{taxosaurus} developed by iPLant and the Phylotastic organization. In this example, we provide a list of species names, some of which are misspelled, and we'll call the API with the \emph{tnrs} function.

<<tnrs, eval=FALSE,cache=TRUE>>=
mynames <- c("Helianthus annuus", "Pinus contort", "Poa anua", "Abis magnifica",
  	"Rosa california", "Festuca arundinace", "Sorbus occidentalos","Madia sateva")
tnrs(query = mynames)[,-c(5:7)]
@

It turns out there are a few corrections: e.g., \emph{Madia sateva} should be \emph{Madia sativa}, and \emph{Rosa california} should be \emph{Rosa californica}. Note that this search worked because fuzzy matching was employed to retrieve names that were close, but not exact matches. Fuzzy matching is only available for plants in the TNRS service, so we advise using EOL's Global Names Resolver if you need to resolve animal names.

@@

<<itisclass, eval=FALSE,cache=TRUE>>=
specieslist <- c("Abies procera","Pinus contorta")
classification(get_tsn(specieslist, "sciname"))
@

@@

Instead of a full classification, you may only want a single name, say a family name for your species of interest. The function \emph{tax\_name} is built just for this purpose. And you can specify the data source you retrieve the taxonomic name from with the \emph{db} parameter, either ITIS or NCBI.

<<taxname, eval=FALSE,cache=TRUE>>=
tax_name(query="Helianthus annuus", get="family", db="itis")
tax_name(query="Helianthus annuus", get="family", db="ncbi")
@

\subsection*{Interactive name selection}
In the previous section we used the function \emph{get\_tsn} to get a classification for two plant species. Below are a few examples. When you run these examples in R, you are presented with a command prompt asking for the row that contains the name you would like back; that output is not printed below for brevity. In this example, the search term has many matches. The function returns a data.frame of the matches, and asks for the user to input what row number to accept.

<<interactive, eval=FALSE, echo=TRUE, cache=TRUE>>=
get_tsn(searchterm="Elisabethae", searchtype="sciname")
@

In another example, you can pass in a long list of taxonomic names:

<<interactive_many, eval=FALSE, echo=TRUE, cache=TRUE>>=
splist <- c("annona cherimola", 'annona muricata', "quercus robur",
  	"shorea robusta", "pandanus patina", "oryza sativa", "durio zibethinus")
get_tsn(searchterm=splist, searchtype="sciname")
@

In another example, note that no match at all returns an NA:

<<interactive_na, eval=FALSE, echo=TRUE, cache=TRUE>>=
get_uid(sciname=c("Chironomus riparius", "aaa vva"))
@

@@

Behind the scenes the function \emph{phylomatic\_tree} retrieves a Taxonomic Serial Number (TSN) from ITIS for each species name, then a string is created for each species like this \emph{poaceae/oryza/oryza\_sativa} (with format 'family/genus/genus\_epithet'). These strings are submitted to the Phylomatic API, and if no errors occur, a phylogeny in newick format is returned. The \emph{phylomatic\_tree()} function also cleans up the newick string and converts it to an \textbf{ape} \emph{phylo} object. The output from \emph{phylomatic\_tree()} is a \emph{phylo} object, which can be used for plotting and phylogenetic analyses. Be aware that Phylomatic has certain limitations - refer to the paper describing Phylomatic \cite{webb2005} and the website \url{http://phylodiversity.net/phylomatic/}.

@@

<<downstream, eval=FALSE, cache=TRUE>>=
col_downstream(name="Apis", downto="Species")[[1]]
@

@@

<<iucn, eval=FALSE, cache=TRUE>>=
ia <- iucn_summary(c("Panthera uncia", "Lynx lynx"))
iucn_status(ia)
@

@@

<<genbank_one, eval=FALSE, cache=TRUE>>=
out <- get_genes_avail(taxon_name="Umbra limi", seqrange = "1:2000", getrelated=FALSE)
@

Then we can ask if 'RAG1' exists in any of the gene names.

<<genbank_two, eval=FALSE, cache=TRUE>>=
out[grep("RAG1", out$genesavail, ignore.case=T),]
@

@@

To analyse trait-composition abundance data must be matched with trait databases like \cite{usseglio-polatera_biological_2000}. However these two data tables may contain species information on different taxonomic levels and perhabs data must be aggregated to a joint taxomic level. Therefore we have to search the taxomomic ladder up- and downwards for matches.

@@

\begin{table}[!ht]
\caption{
\bf{Data sources used in taxize, tasks available, and links to them}}
\begin{tabular}{|l|c|c|c|c|}
\hline
Source name & Name search & Name resolution & Phylogeny & Sequences  \\
\hline
Encyclopedia of Life \cite{eol} & Yes & See GNR below & No & No \\
Integrated Taxonomic Information System \cite{itis} & Yes & Synonyms & No & No \\
iPlant Taxonomic Name Resolution Service \cite{tnrs} & Yes & Yes & No & No \\
Phylomatic \cite{webb2005} & No & No & Yes & No \\
uBio \cite{ubio} & Yes & Yes & No & No \\
Global Names Resolver \cite{eolgnr} & Yes & Yes & No & No \\
Global Names Index \cite{eolgnr} & Yes & No & No & No \\
IUCN Red List \cite{iucn} & Yes & No & No & No \\
Tropicos \cite{tropicos} & Yes & Yes & No & No \\
Plantminer \cite{carvalho2010plantminer} & Yes & No & No & No \\
The Plant List & Yes & Yes & No & No \\
Catalogue of Life \cite{col} & Yes & Yes & No & No \\
National Center for Biotechnology Information \cite{federhen} & Yes & X & Yes \textsuperscript{a} & Yes \\
\hline
\end{tabular}
\begin{flushleft}\textsuperscript{a} Web only, no API, see \url{http://1.usa.gov/11R446a}
\end{flushleft}
\label{tab:a}
\end{table}
